---
title: "Machine Learning Wines"
author: "Ron Segal"
date: "16/08/2015"
output: html_document
runtime: shiny
---

## Introduction

This R Shiny application was developed as part of an assignment for the Coursera 'Developing Data Products' data science series course. It uses the k-means clustering, unsupervised, machine learning algorithm, to try to recognise which wines are attributable to which of 3 growers from chemical analysis data.  The data set was downloaded from the UCI machine learning repository (archive.ics.uci.edu/ml/datasets/Wine). It is for 178 instances of wines grown in the same region of Italy, providing the quantities of 13 constituents found in each of three types of wine.

The objective of the application is to graphically explore and try to identify the number of k-means clusters (which must be specified in advance of running the algorithm) that best partitions the data, assuming that we don't know in advance that there are three different types of wine. Effectively this is a kind of machine wine "tasting" to try to recognise wines coming from different (anonymous) growers!

The k-means clusters algorithm operates upon all 13 variables, with the wine "type" variable being dropped from the set. The output is projected onto a two dimensional scatterplot by using discrproj (from R package fpc), which implements linear dimension reduction. In addition the Silhouette method (ftp://adrem.ua.ac.be/pub/preprints/87/Silgra87.pdf) is used to calculate an average silhouette width, which provides an evaluation of clustering validity, which can be used to select an 'appropriate' number of clusters.  (a silhouette plot isn't shown).

## Usage

The number of clusters to be used by the k-means clustering algorithm is selected using the slider on the left of the plot. When the slider is moved the plot immediately shows the result, with different colours representing each cluster.

At the bottom of the plot there is a red band with a numeric label indicating the average silhouette width. (Note that the silhouette algorithm isn't valid for a single cluster, so the band isn't shown in that case.)  A larger average width is indicative of a more appropriate number of clusters.

A 'Show Wine Growers' check box overlays the plot with wine 'type' labels, which indicate the grower to which each wine sample is attributable.

Finally, a numeric input box enables the value of the pseudo random number 'seed' to be set, which changes the sequence of random numbers used in the calculations, which for example randomises the start centroid of each cluster.


``` {r, echo = FALSE, message=FALSE}

# clear global environment
# rm(list=ls())
library(shiny)
library(cluster)
library(HSAUR)
library(rainbow)
library(ggplot2)
library(fpc)

winedata<<-read.csv("data/wine.data", sep=",")
# Wine data from: http://archive.ics.uci.edu/ml/datasets/Wine

names(winedata)<<- c("Type","Alcohol","Malic", "Ash", "Alcalinity", "Magnesium", "Phenols", "Flavanoids", "Nonflavanoids", "Proanthocyanins", "Colorintensity", "Hue", "Dilution", "Proline")           

wines<<-winedata
wines$Type<-NULL    # Remove the 'Type' variable for 3 growers, i.e. values 1,2, or 3
wines<-scale(wines) # Normalise the scale of the independent variables

# Function generates a ramp of color values from a vector
cRamp <<- function(x){
  if (diff(range(x)) > 0) {
    x <- (x-min(x))/diff(range(x))
  }
  cols <- colorRamp(rainbow(9))(x)
  apply(cols, 1, function(xt)rgb(xt[1], xt[2], xt[3], maxColorValue=255))
}  
  
```


## Embedded Application

It's also possible to embed an entire Shiny application within an R Markdown document using the `shinyAppDir` function. This example embeds a Shiny application located in another directory:

```{r, echo = FALSE, message=FALSE}

shinyApp(
  
  ui = pageWithSidebar(
  headerPanel(HTML('<p style="color:darkred; font-size:30px">Which Wine Grower?</p>')),
  sidebarPanel(

    sliderInput("clusters", "K-means Cluster count:", 
                min=1, max=8, value=1),
    br(),
    br(),
    checkboxInput(inputId = "growers",
      label = strong("Show Wine Growers"),
      value = FALSE),
      br(),
      br(),
    numericInput("seedval", "Change Seed Value:", 
                value=999)      
   ),
   
  mainPanel(
    plotOutput('plot1')
  )
),
  

  server = function(input, output, session) {

  totalClusters <- reactive({
       set.seed(input$seedval) # Get seed value from numeric input
       # Run k-means model
       km <<- kmeans(wines,input$clusters)
       # Run silhouette model to identify number of clusters
       # that achieve best fit as indicated by average width
       dissE <- daisy(wines) 
       dE2   <- dissE^2
       sk2   <- silhouette(km$cl, dE2)
       # Silhouette method is only valid for > 1 clusters
       if (input$clusters > 1) {
        sk2avwidth<<-summary(sk2)$avg.width
       } else {
        sk2avwidth<<-0 # value used to indicate an error
       }
       # Note - discrproj (from fpc) implements linear dimension reduction for classification
       # $proj contains coordinates of the original data points projected onto a 2d space
       # It has the same dimensions as the original with the first two dimensions
       # providing the best separation of the clusters
       
       xprojection<-discrproj(wines, km$cluster)$proj # See note above 
       xprojframe<<-data.frame(xprojection)
       colMin <<- sapply(xprojframe, min, na.rm = TRUE) # Get minimum column values plot positioning
       
       input$clusters # return the number of clusters from slider
  })
  

  output$plot1 <- renderPlot({
    numclusters<-totalClusters()
    par(mar = c(6.1, 5.1, 0, 1))
    # Show grower designators if check box ticked
    if (input$growers) { 
    plotgrowers <- geom_text(data=xprojframe, mapping=aes(x=X1, y=X2, label=winedata$Type), size=4,  color="black")
    } else {
    plotgrowers <- NULL}
    
    # Plot silhouette average width (scaled up) with value as label
    s1 <- geom_segment(aes(colMin["X1"],colMin["X2"]-1, xend=sk2avwidth*10, yend=colMin["X2"]-1),color="red",size=8, alpha=0.5)
    s2 <- geom_text(aes(x=sk2avwidth*10,y=colMin["X2"]-1, label=round(sk2avwidth,digits=2),hjust=1))
    s3 <- geom_text(aes(x=sk2avwidth*10,y=colMin["X2"]-1, label="Silhouette Av Width", vjust=-1, hjust=1))
    s4 <- geom_text(aes(x=colMin["X1"],y=colMin["X2"]-0.5, label="Silhouette Not Applicable with 1 Cluster", hjust=0))
    # A width of 0 is used to indicate an error, generally when number of clusters < 2
    if (sk2avwidth > 0) {
    plotsilhouette <- c(s1,s2,s3)
    } else {
    plotsilhouette <- s4}
    
    ggplot() +
    # Scatterplot of k-means model
    geom_point(data=xprojframe, mapping=aes(x=X1, y=X2, fill=cRamp(km$cluster)), size=6, shape=21, color="black") +
    ggtitle("Machine Wine \"Tasting\" Clusters") +
        theme(plot.title = element_text(size=22, face="bold", colour="darkred")) +
      # Remove axes ticks and text
      theme(axis.ticks.x=element_blank(), axis.ticks.y=element_blank(),axis.text.x=element_blank(), axis.text.y=element_blank(),axis.title.x=element_blank(), axis.title.y=element_blank()) +
        # Modify legend
        scale_fill_discrete(name="Clusters", labels=seq(1,numclusters,1)) +
            plotgrowers +
                plotsilhouette
                
  })
  
  
},

options = list(height=450)

)
```




