---
title: "Machine Learning Wines"
author: "Ron Segal"
date: "16/08/2015"
output: html_document
runtime: shiny
---

## Introduction

This R Shiny application was developed as part of an assignment for the Coursera 'Developing Data Products' data science series course. It uses the k-means clustering, unsupervised, machine learning algorithm, to try to recognise from chemical analysis data, which wines are attributable to which of 3 growers.  The data set was downloaded from the UCI machine learning repository (http://archive.ics.uci.edu/ml/datasets/Wine); it provides the quantities of 13 constituents found in 178 instances of wine coming from 3 different growers in the same region of Italy.

The objective of the application is to graphically explore and try to identify the number of k-means clusters that best partitions the wine data. Effectively this is a kind of machine wine "tasting" to try to separate wines coming from different (anonymous) cultivators!

The k-means clusters algorithm operates upon all 13 variables, with the wine "type" variable being removed from the set for the purposes of the analysis. The output is projected onto a two dimensional scatterplot, using discrproj (from the R package fpc), which implements linear dimension reduction. In addition the Silhouette method (ftp://adrem.ua.ac.be/pub/preprints/87/Silgra87.pdf) is used to calculate an average silhouette width, which provides an evaluation of clustering validity, which can be used to select an 'appropriate' number of clusters.  (a silhouette plot itself isn't shown).

## Usage

The number of clusters to be used by the k-means clustering algorithm is selected using the slider on the left of the plot. When the slider is moved the plot immediately shows the resulting clusters, with different colours representing each cluster.

At the bottom of the plot there is a red band with a numeric label indicating the average silhouette width. (Note that the silhouette algorithm isn't valid for a single cluster, so the band isn't shown in that case.)  A larger average width is indicative of better cluster separation.

Checking the 'Show Wine Growers'check box overlays the plot with wine 'type' numeric labels, which indicate the grower to which each wine sample is in reality attributable. From this it can be seen how accurately the clusters are allocating wines to growers.  Clearly this will largely depend on how much variance in the chemical quantities is attributable to different growers. It is also possible of course that different conditions within a grower could giver rise to marked differences in chemical composition, which would likely be reflected in the clustering.

Finally, a numeric input box enables the value of the pseudo random number 'seed' to be modified. This changes the sequence of random numbers used in the calculations, which determines the initial starting position of each cluster, which the k-means algorithm tries to optimise.


``` {r, echo = FALSE, message=FALSE, warning=FALSE}

options(rgl.useNULL=TRUE) #fixes a warning message
library(shiny)
library(cluster)
library(HSAUR)
library(rainbow)
library(ggplot2)
library(fpc)

winedata<<-read.csv("data/wine.data", sep=",")
# Wine data from: http://archive.ics.uci.edu/ml/datasets/Wine

# Most of these variable names aren't actually used in the code below
names(winedata)<<- c("Type","Alcohol","Malic", "Ash", "Alcalinity", "Magnesium", "Phenols", "Flavanoids", "Nonflavanoids", "Proanthocyanins", "Colorintensity", "Hue", "Dilution", "Proline")  

wines<<-winedata
wines$Type<-NULL    # Remove the 'Type' variable for 3 growers, i.e. values 1,2, or 3
wines<-scale(wines) # Normalise the scale of the independent variables

# Function generates a ramp of color values from a vector
cRamp <<- function(x){
  if (diff(range(x)) > 0) {
    x <- (x-min(x))/diff(range(x))
  }
  cols <- colorRamp(rainbow(9))(x)
  apply(cols, 1, function(xt)rgb(xt[1], xt[2], xt[3], maxColorValue=255))
}  
  
```


```{r, echo = FALSE, message=FALSE, warning=FALSE}

shinyApp(
  
  ui = pageWithSidebar(
  headerPanel(HTML('<p style="color:darkred; font-size:30px">Which Wine Grower?</p>')),
  sidebarPanel(

    sliderInput("clusters", "K-means Cluster count:", 
                min=1, max=8, value=1),
    br(),
    br(),
    checkboxInput(inputId = "growers",
      label = strong("Show Wine Growers"),
      value = FALSE),
      br(),
      br(),
    numericInput("seedval", "Change Seed Value:", 
                value=999)      
   ),
   
  mainPanel(
    plotOutput('plot1')
  )
),
  

  server = function(input, output, session) {

  totalClusters <- reactive({
       set.seed(input$seedval) # Get seed value from numeric input
       # Run k-means model
       km <<- kmeans(wines,input$clusters)
       # Run silhouette model to identify number of clusters
       # that achieve best fit as indicated by average width
       dissE <- daisy(wines) 
       dE2   <- dissE^2
       sk2   <- silhouette(km$cl, dE2)
       # Silhouette method is only valid for > 1 clusters
       if (input$clusters > 1) {
        sk2avwidth<<-summary(sk2)$avg.width
       } else {
        sk2avwidth<<-0 # value used to indicate an error
       }
  # Note - discrproj (from fpc) implements linear dimension reduction for classification
  # $proj contains coordinates of the original data points projected onto a 2d space
  # It has the same dimensions as the original with the first two dimensions
  # providing the best separation of the clusters
       
       xprojection<-discrproj(wines, km$cluster)$proj # See note above 
       xprojframe<<-data.frame(xprojection)
       colMin <<- sapply(xprojframe, min, na.rm = TRUE) # Get minimum column values plot positioning
       
       input$clusters # return the number of clusters from slider
  })
  

  output$plot1 <- renderPlot({
    numclusters<-totalClusters()
    par(mar = c(6.1, 5.1, 0, 1))
    
    # Show grower designators if check box ticked
    if (input$growers) { 
    plotgrowers <- geom_text(data=xprojframe, mapping=aes(x=X1, y=X2, label=winedata$Type), size=4,  color="black")
    } else {
    plotgrowers <- NULL}
    
    # Plot silhouette average width (scaled up) with value as label
    s1 <- geom_segment(aes(colMin["X1"],colMin["X2"]-1, xend=sk2avwidth*10, yend=colMin["X2"]-1),color="red",size=8, alpha=0.5)
    s2 <- geom_text(aes(x=sk2avwidth*10,y=colMin["X2"]-1, label=round(sk2avwidth,digits=2),hjust=1))
    s3 <- geom_text(aes(x=sk2avwidth*10,y=colMin["X2"]-1, label="Silhouette Av Width", vjust=-1, hjust=1))
    s4 <- geom_text(aes(x=colMin["X1"],y=colMin["X2"]-0.5, label="Silhouette Not Applicable with 1 Cluster", hjust=0))

# A width of 0 is used to indicate an error, generally when number of clusters < 2
    if (sk2avwidth > 0) { # Silhouette is valid so show
    plotsilhouette <- c(s1,s2,s3)
    } else {
    plotsilhouette <- s4} # Silhouette not valid
    
    ggplot() +
    # Scatterplot of k-means model
    geom_point(data=xprojframe, mapping=aes(x=X1, y=X2, fill=cRamp(km$cluster)),
               size=6, shape=21, color="black") +
    ggtitle("Machine Wine \"Tasting\" Clusters") +
        theme(plot.title = element_text(size=22, face="bold", colour="darkred")) +
      # Remove axes ticks and text
      theme(axis.ticks.x=element_blank(),
            axis.ticks.y=element_blank(),axis.text.x=element_blank(),
            axis.text.y=element_blank(),axis.title.x=element_blank(),
            axis.title.y=element_blank()) +
        # Modify legend
        scale_fill_discrete(name="Clusters", labels=seq(1,numclusters,1)) +
            plotgrowers +
                plotsilhouette
                
  })
  
  
},

options = list(height=500) # The height of the plot

)
```




